Estimating SARS-CoV-2 transmission in England from randomly sampled prevalence surveys

### 2024-02-09 transmission.tex
Overleaf comments

A parsimonious model is needed to parameterise the WAIFW matrix (see \cref{E-SEIR:sec:structured-populations}).
I follow the WAIFW matrix parameterisation of \textcite{birrellRealtime}, decomposing the rate of transmission from infectious individuals in age group $a'$ to susceptible individuals in age group $a$ as $\beta_{aa'}(t) = \beta(t) \kappa_{aa'}(t) q_{aa'} N_{a'}$.
This separates contact patterns, $\kappa_{aa'}(t)$, which can vary over time; probabilities of transmission upon contact between age groups $a$ and $a'$, $q_{aa'}$, assumed constant; a region-wide effect for time, $\beta(t)$; and the population of the stratum, $N_{a'}$.

@@

Some behavioural changes (\eg mask wearing) impact transmission.
However, these are not captured by the contact survey.
To allow for these changes and the uncertainty produced in formulating $\matr{\kappa}(t)$, I follow \textcite{birrellRealtime} in placing a random walk prior on $\log\beta(t)$  on the weekly timescale:
\begin{align}
    \beta(t) &= \beta_{\tilde{w}(t)} \\
    \log\beta_{w+1} &= \log\beta_w + \sigma_\epsilon \epsilon_w \\
    \epsilon_w &\sim \N(0, 1)
\end{align}
where $\tilde{w}(t)$ is the week number of time $t$, with week 1 being the first week considered.
Two hyperparameters will be inferred: $\beta_1$, the initial value of the random walk; and $\sigma_\epsilon$, the standard deviation of the random walk (see \cref{SEIR:sec:priors}).

@@

Finally, $q_{aa'}$ needs to be specified.
I follow a common approach\todo{cite anderson and may}, decomposing it into two independent factors such that $q_{aa'} = q^\text{susc}_{a} q^\text{inf}_{a'}$, where $q^\text{susc}_a$ is the \emph{susceptibility} of stratum $a$, \ie how vulnerable to infection individuals in stratum $a$ are; and $q^\text{inf}_{a'}$ is the \emph{transmissibility} of stratum $a'$, \ie how likely individuals in stratum $a$ are to pass on an infection.
% Then $\beta^\text{susc}_i$ is the relative risk of an individual in stratum $i$ being infected, relative to an individual in stratum $a_0$.
These measures include both biological differences, \eg due to differences in immune systems, or behavioural, \eg if infants are less hygienic than adults.

@@

A computationally cheaper option, which I will take, is to approximate this convolution by adding compartments to the ODEs.
This approach allows an infected, pre-positive compartment, which I denote $p_0$, to be included.
$p_0$ increases at the same rate as individuals flow into $e_1$.
These individuals then transition into a set of compartments with transition rates, $p_1, \dots, p_l$, chosen to match the duration of positivity estimates in \cref{E-imperf-test}.
This will lead to the model in FIGURE - with transition rates etc etc
The new compartments do not have a biological interpretation; the same idea is behind having multiple exposed or latent compartments (introduced in \cref{E-SEIR:sec:non-exponential}).

Deciding $l$ and the length of time each individual spends in each compartment is the most challenging part of formulating the observation model.
The class of distributions which can be represented as a series of compartments are known as \emph{phase-type} distributions.
Discussions of general phase-type distributions are beyond the scope of this thesis; a review in a biological context can be found in \textcite{hobolthPhasetype}.
I will restrict myself to the subset of phase-type distributions defined by \textcite{osogamiClosed} as Erlang-Coxian (EC) distributions.
EC distributions are convenient because they are flexible enough to approximate almost any distribution, and the parameters of the approximation are given by closed-form functions of the distribution's moments.

% Formally, a phase-type distribution as the distribution of \emph{absorption time} in a \emph{continuous-time Markov process} with a single absorbing state~\autocite{hobolthPhasetype}.
% Given a probability distribution over starting states, a rate at which 
% The process can be defined by its size, $l$, a $l \times l$ \emph{transition matrix}, $\matr{M} \in \reals^{l \times l}$, an \emph{exit rate} vector, $\vec{\alpha}$ 
% The process's state at time $t$, $W_t$, is a scalar $W_t \in \{ 1, \dots, l+1 \}$.
% Its transition matrix is defined, for $j \neq k$, as $\prob(W_{t+\delta t} = k \mid ) = M_{jk} \delta t$ for small, positive $\delta t$.
% The diagonal is defined as $M_{jj} = - \sum_{k \neq j} M_{jk}$ so that the matrix $\matr{M}$ has row sums 0.
% The state $l+1$ is an absorbing state, that is $\prob(W_{t+t'} = l+1 \mid W_t = l_1) = 1$ for any non-negative $t'$ (once the process enters it, it cannot leave).
% The exit rate gives the transition rates into the absorbing state: $\prob(W_{t+\delta t} = l + 1 \mid W_t = j) = \alpha_j \delta t$ for small $\delta t$.
% The process's absorption time is when it first arrives at the absorbing state, $\tau = \inf \{ t \ssep W_t = l+1 \}$.
% That is, $P$ is a phase-type distribution if it can be represented as a process 
% They are additions of arbitrary exponential distributions, and can (if there is no limit on the number of compartments used) represent any distribution.
% \todo{check/cite these claims}
% I use the method of \textcite{osogamiClosed} to determine $l$ and the transition rates between the compartments.

% A $l$-phase EC distribution is the sum of a $(l-2)$-phase gamma distribution and a two-phase acyclic, arbitrary phase-type distribution.
An $l$-phase EC distribution $P$ can be defined as the sum of three random variables: $K$, $X_1$, and $X_2$.
It is shown schematically in \cref{SEIR:fig:EC}.
Here, $K \dist \GamDist(l-2, \omega_Y)$, $X_1 \dist \Exponential(\omega_{X1})$ and $X_2$ is a mixture distribution taking the value 0 with probability $1-p_x$ or otherwise (\ie with probability $p_x$) distributed $\Exponential(\omega_{X2})$.
% The time between entering the 1st state and the $l+1$th state, the absorbing state, is the duration.
These distributions contain five parameters.
Three are positive reals, $\omega_Y$, $\omega_{X1}$, and $\omega_{X2}$; one is a probability, $p_x$, in the interval $[0, 1]$; and one is a positive integer, $l$.

% The name Erlang-Coxian comes from the two distributions that make up the EC distribution.
% Gamma distributions with integer first parameters (as $K$ has) are also known as Erlang distributions.
% The distribution $X_1 + X_2$ is a special case of the Coxian distribution.
% Hence, the overall distribution is known as a Erlang-Coxian distribution.

    \node[state] (1) {1};
    \node[state, right of=1] (2) {2};
    \node[state, right of=2, draw=none] (dots) {$\dots$};
    \node[state, right of=dots] (lm2) {$l-2$};
    \node[state, right of=lm2] (lm1) {$l-1$};
    \node[state, right of=lm1] (l) {$l$};
    \node[state, right of=l, align=center] (absorbing) {absorbing\\state};

    \path (1) edge node {$\omega_Y$} (2)
          (2) edge node {$\omega_Y$} (dots)
          (dots) edge node {$\omega_Y$} (lm2)
          (lm2) edge node {$\omega_Y$} (lm1)
          (lm1) edge node {$p_x \omega_{X1}$} (l)
                edge [out=-45,in=-135] node[below] {$(1 - p_x) \omega_{X1}$} (absorbing)
          (l) edge node {$\omega_{X2}$} (absorbing);
\end{tikzpicture}
}
\caption[A $l$ phase EC distribution.]{Representation of an EC distribution as a series of compartments. The distribution is the time from entering state 1 and entering the absorbing state. The time between entering state 1 and state $l-2$ is distributed $K \dist \GamDist(l-2, \omega_Y)$, also known as an Erlang distribution. The time between entering state $l-2$ and the absorbing state has a Coxian distribution (see the main text for details).}
\label{SEIR:fig:EC}
\end{figure}

\Textcite{osogamiClosed} define a phase-type distribution $P$ as well representing an arbitrary distribution $G$ if $P$ and $G$ have the same first three moments.
They then prove that EC distributions have the following properties.
\begin{itemize}
    \item Let $\set{W}$ be the set of all phase-types distributions that well represent $G$ and $w$ be the minimum number of phases of any member of $\set{W}$. If $\set{W}$ is non-empty, then there exists a $l$-phase EC distribution $P' \in \set{W}$.
    \item Furthermore, at least one such $P'$ has $l \leq w + 1$. Fewer phases is generally beneficial as it reduces the time and space complexity of working with the distribution.
    \item The parameters of a $P'$ with $l \leq w + 1$ can be found using a closed-form expression (given by \textcite{osogamiClosed}). This is beneficial because finding arbitrary members of $\set{W}$ is a hard problem.
\end{itemize}

A total of five parameters are required to specify a $l$-phase EC distribution.
\begin{itemize}
    \item $l$, the number of phases.
    \item $\omega_Y$, the rate parameter of the Erlang distribution.
    \item $\omega_{X1}$, the rate parameter of the first state after the Erlang distribution.
    \item $\omega_{X2}$, the rate parameter of the second state after the Erlang distribution.
    \item $p_x$, the probability of moving from the end state of the Erlang distribution to the second (as opposed to directly to the absorbing state).
\end{itemize}

The mapfit package~\autocite{mapfit} implements the expressions to calculate the EC distribution parameters given a distribution's first three moments.
Applying the package to the posterior mean of the distribution estimated in \cref{E-imperf-test} gives the parameter estimates in \cref{SEIR:table:ec-params}.
\begin{table}
    \centering
    \begin{tabular}{c c c c c}
        $l$ & $\omega_Y$ & $\omega_{X1}$ & $\omega_{X2}$ & $p_x$ \\
        3 & 0.0820 & 0.126 & 0.0223 & 0.00973  \\
    \end{tabular}
    \caption{Parameter estimates for the EC distribution approximating the posterior mean of the duration distribution estimated in \cref{E-imperf-test}.}
    \label{SEIR:table:ec-params}
\end{table}
% end table of ec params

I link the modelled proportion of individuals that are RT-PCR positive to the data using a beta-binomial likelihood (as discussed in \cref{transmission:sec:data}).
If on each day $t$, each strata $i$ is observed to have $y_{it}$ positives out of $n_{it}$ tests, then the likelihood is:
\begin{align}
    \prod_{a,t} \BB (y_{at} \mid n_{at}, \sum_{k=1}^l \bar{p}_{kat}, \rho)
    \label{SEIR:eq:likelihood}
\end{align}
where $\bar{p}_{kat} = \int_{t}^{t+1} p_{ka}(s) ds$ is the mean occupancy in compartment $p_k$ and strata $a$ over day $t$ and $\rho$ is a nuisance parameter controlling the overdispersion of the beta-binomial distribution.

@@

    \path  (S) edge node {$\lambda$} (P0)
          (P0) edge node {$\omega_0$} (P1)
          (P1) edge node {$\omega_Y$} (P2)
          (P2) edge node {$p_x \omega_{X1}$} (P3)
               edge [out=-45,in=-135] node[below] {$(1 - p_x) \omega_{X1}$} (P4)
          (P3) edge node {$\omega_{X2}$} (P4);

@@

A diagram of the model for a single stratum is shown in figure \cref{SEIR:fig:full-model}.
The full system of differential equations for each region is below.
\begin{align}
    \label{SEIR:eq:fullODEs}
    \vec{\lambda}(t) &= \beta(t) (\matr{\kappa}(t) \circ \matr{q}) (\vec{i_1}(t) + \vec{i_2}(t)) \\
    \frac{d\vec{s}(t)}{dt} &= -\vec{\lambda}(t) \circ \vec{s}(t) \\
    \frac{d\vec{e_1}(t)}{dt} &= \vec{\lambda}(t) \circ \vec{s}(t) - 2\sigma \vec{e_1}(t) \\
    \frac{d\vec{e_2}(t)}{dt} &= 2\sigma \vec{e_1}(t) - 2\sigma \vec{e_2}(t) \\
    \frac{d\vec{i_1}(t)}{dt} &= 2\sigma \vec{e_2}(t) - 2\gamma \vec{i_1}(t) \\
    \frac{d\vec{i_2}(t)}{dt} &= 2\gamma \vec{i_1}(t) - 2\gamma \vec{i_2}(t) \\
    \frac{d\vec{r}(t)}{dt} &= 2\gamma \vec{i_2}(t) \\
    \frac{d\vec{p_0}(t)}{dt} &= \vec{\lambda} \circ \vec{s}(t) - \omega_0 \vec{p_0}(t) \\
    \frac{d\vec{p_1}(t)}{dt} &= \omega_0 \vec{p_0}(t) - \omega_Y \vec{p_1}(t) \\
    \frac{d\vec{p_2}(t)}{dt} &= \omega_Y \vec{p_1}(t) - \omega_{X1} \vec{p_2}(t) \\
    \frac{d\vec{p_3}(t)}{dt} &= p_x \omega_{X1} \vec{p_2}(t) - \omega_{X2} \vec{p_3}(t).
\end{align}

@@

The mean prevalence on day $t$, used for the likelihood (see \cref{SEIR:eq:likelihood}), can be calculated as $\bar{p}_{kat} = \frac{1}{2} ( p_{kat} + p_{ka,t+1/2} )$.

@@

From the system of ODEs, the knowledge that all the compartments are growing exponentially at rate $\psi$, and the parameters $\vec{\pi}$ (where $\pi_a$ is the proportion of strata $i$ without immunity at time 0, hence, $\vec{r} = 1 - \vec{\pi}$) one can derive the following for the initial state of the compartments:
\begin{align}
    \vec{i_1} &= \vec{i_0} \left(1 + \frac{2\gamma}{\psi + 2\gamma} \right)^{-1} \\
    \vec{i_2} &= \vec{i_0} - \vec{i_1} \\
    \vec{e_2} &= \vec{i_1} \frac{\psi + 2\gamma}{2\sigma} \\
    \vec{e_1} &= \vec{e_2} \frac{\psi + 2\sigma}{2\sigma} \\
    \vec{s} &= \vec{\pi} - \vec{i_0} - \vec{e_1} - \vec{e_2} \\
    \vec{p_0} &= \vec{e_1} \frac{\psi + 2\sigma}{\psi + \omega_0} \\
    \vec{p_1} &= \vec{p_0} \frac{\omega_0}{\psi + \omega_Y} \\
    \vec{p_2} &= \vec{p_1} \frac{\omega_Y}{\psi + \omega_{X1}} \\
    \vec{p_3} &= \vec{p_2} \frac{p_x \omega_{X1}}{\psi + \omega_{X2}}.
\end{align}

@@

\ifSubfilesClassLoaded{
  \listoftodos
}{}

>>>>

A parsimonious model is needed to parameterise the WAIFW matrix.
In \cref{E-SEIR:sec:structured-populations}, I explained how the elements of the WAIFW matrix, $\matr{\beta}$ can be decomposed as $\beta_{aa'} = \kappa_{aa'} q_{aa'} N_{a'}$ where\todo{should I repeat these definitions? they already exist in ch 3}: $\kappa_{aa'}$ is the mean rate at which an individual in stratum $a'$ makes contact with individuals in stratum $a$; $q_{aa'}$ is the probability of transmission upon contact between an infectious individual in stratum $a'$ and a susceptible individual in stratum $a$; and $N_{a'}$ is the total number of individuals in stratum $a'$.
For many epidemics, where the population behaviour is fairly constant over time, a constant $\matr{\beta}$ is a good approximation.
However, in SARS-CoV-2, population behaviour changed dramatically, both spontaneously (\eg due to fear) and in response to government-imposed restrictions.
Therefore, following \textcite{birrellRealtime}, I introduce a time-dependence giving: $\beta_{aa'}(t) = \kappa_{aa'}(t) q_{aa'}(t) N_{a'}$; $N_{a'}$ is constant because the change in the total population is negligible over the length of time considered here.

@@

The probability of transmission, $q_{aa'}(t)$, needs further decomposition to be identifiable, for the same issue of dimensionality preventing each element of $\matr{\beta}$ being estimated independently (see \cref{SEIR:sec:structured-populations} for a full explanation).
I assume that $q_{aa'}(t) = \beta(t) q^\text{susc}_{a} q^\text{inf}_{a'}$ where $\beta(t)$ is a time-varying reference probability of transmission; $q^\text{susc}_a$ is the relative \emph{susceptibility} of stratum $a$, \ie how vulnerable to infection individuals in stratum $a$ are; and $q^\text{inf}_{a'}$ is the relative \emph{transmissibility} of stratum $a'$, \ie how likely individuals in stratum $a$ are to pass on an infection.
Separating relative transmissibility and susceptibility into two independent factors is a common approach\todo{cite anderson and may}.
These measures include both biological differences, \eg due to differences in immune systems, or behavioural, \eg if infants are less hygienic than adults.

The time-varying component, $\beta(t)$, allows for behavioural changes that impact transmission but are not captured by the contact matrices (\eg mask wearing).
I follow prior work~\autocite[e.g.][]{birrellRealtime}\todo{find other references for random walk on transmission potential} in using a random walk prior on $\log\beta(t)$  on the weekly timescale.
This prior encodes that behaviour changes smoothly over time.
\begin{align}
    \beta(t) &= \beta_{\tilde{w}(t)} \\
    \log\beta_{w+1} &= \log\beta_w + \sigma_\epsilon \epsilon_w \\
    \epsilon_w &\sim \N(0, 1)
\end{align}
where $\tilde{w}(t)$ is the week number of time $t$, with week 1 being the first week considered.
Two hyperparameters will be inferred: $\beta_1$, the initial value of the random walk; and $\sigma_\epsilon$, the standard deviation of the random walk (see \cref{SEIR:sec:priors}).
$\beta(t)$ is the only time-varying part of the transmission model that is not fixed a priori.
Therefore, in practice, it will also absorb other effects such as any error in $\matr{\kappa}$.

@@

@@

A computationally cheaper option, which I will take, is to approximate this convolution by adding compartments to the ODEs.
This approach allows an infected, pre-positive compartment, which I denote $p_0$, to be included.
Up until now, this stage of the disease (shown in \cref{E-biology-data:fig:natural-history}) has been neglected, assumed to be of negligble duration.
An advantage of this modelling approach is that it can be included without complications (\eg issues of numerical stability).

$p_0$ increases at the same rate as individuals flow into $e_1$.
These individuals then transition into a set of compartments, $p_1, \dots, p_l$, each with their own transition rate.
I will choose $l$ and the transition rates such that combined length of stay (\ie time from entry in $p_1$ until exit from $p_l$) approximates the duration of positivity distribution estimated in \cref{E-imperf-test}.
This idea extends the technique of having multiple exposed or latent compartments (introduced in \cref{E-SEIR:sec:non-exponential}), and is commonly used to enforce correct durations in compartmental models\todo{cite phase-type distributions in compartmental models}.

\Textcite{osogamiClosed} provide a method to construct the compartments, implemented in the mapfit R package~\autocite{mapfit} implements the expressions to calculate the EC distribution parameters given a distribution's first three moments.
Applying the package to the posterior mean of the distribution estimated in \cref{E-imperf-test} gives the model in \cref{SEIR:fig:full-model} with the parameter estimates in \cref{SEIR:table:ec-params} (see \cref{transmission:sec:phase-type} for details).
\begin{table}
    \centering
    \begin{tabular}{c c c c c}
        $l$ & $\omega_Y$ & $\omega_{X1}$ & $\omega_{X2}$ & $\pi$ \\
        3 & 0.0820 & 0.126 & 0.0223 & 0.00973  \\
    \end{tabular}
    \caption{Parameter estimates for the EC distribution approximating the posterior mean of the duration distribution estimated in \cref{E-imperf-test}.}
    \label{SEIR:table:ec-params}
\end{table}

The mapfit R package~\autocite{mapfit} implements the expressions to calculate the EC distribution parameters given a distribution's first three moments.
Applying the package to the posterior mean of the distribution estimated in \cref{E-imperf-test} gives the parameter estimates in \cref{SEIR:table:ec-params}.
\begin{table}
    \centering
    \begin{tabular}{c c c c c}
        $l$ & $\omega_Y$ & $\omega_{X1}$ & $\omega_{X2}$ & $\pi$ \\
        3 & 0.0820 & 0.126 & 0.0223 & 0.00973  \\
    \end{tabular}
    \caption{Parameter estimates for the EC distribution approximating the posterior mean of the duration distribution estimated in \cref{E-imperf-test}.}
    \label{SEIR:table:ec-params}
\end{table}
% end table of ec params

I link the modelled proportion of individuals that are RT-PCR positive to the data using a beta-binomial likelihood (as discussed in \cref{transmission:sec:data}).
If on each day $t$, each strata $i$ is observed to have $y_{at}$ positives out of $n_{at}$ tests, then the likelihood is:
\begin{align}
    \prod_{a,t} \BB (y_{at} \mid n_{at}, \bar{p}_{at}, \rho)
    \label{SEIR:eq:likelihood}
\end{align}
where: $\bar{p}_{at}$ is the mean proportion of individuals in strata $a$ that are RT-PCR positive on day $t$; and $\rho$ is a nuisance parameter controlling the overdispersion of the beta-binomial distribution.
Denoting the proportion of stratum $a$ in state $p_k$ as $p_{ka}(\tau)$, then $\bar{p}_{at} = \int_{t}^{t+1} \sum_{k=1}^l p_{ka}(\tau) d\tau$.

@@

    \path  (S) edge node {$\lambda$} (P0)
          (P0) edge node {$\omega_0$} (P1)
          (P1) edge node {$\omega_Y$} (P2)
          (P2) edge node {$\pi \omega_{X1}$} (P3)
               edge [out=-45,in=-135] node[below] {$(1 - \pi) \omega_{X1}$} (P4)
          (P3) edge node {$\omega_{X2}$} (P4);

@@

A diagram of the model for a single stratum is shown in figure \cref{SEIR:fig:full-model}.
The full system of differential equations for each region is below.
\begin{align}
    \label{SEIR:eq:fullODEs}
    \vec{\lambda}(t) &= \beta(t) (\matr{\kappa}(t) \circ \matr{q}) ((\vec{i_1}(t) + \vec{i_2}(t)) \circ \vec{N}) \\
    \frac{d\vec{s}(t)}{dt} &= -\vec{\lambda}(t) \circ \vec{s}(t) \\
    \frac{d\vec{e_1}(t)}{dt} &= \vec{\lambda}(t) \circ \vec{s}(t) - 2\sigma \vec{e_1}(t) \\
    \frac{d\vec{e_2}(t)}{dt} &= 2\sigma \vec{e_1}(t) - 2\sigma \vec{e_2}(t) \\
    \frac{d\vec{i_1}(t)}{dt} &= 2\sigma \vec{e_2}(t) - 2\gamma \vec{i_1}(t) \\
    \frac{d\vec{i_2}(t)}{dt} &= 2\gamma \vec{i_1}(t) - 2\gamma \vec{i_2}(t) \\
    \frac{d\vec{r}(t)}{dt} &= 2\gamma \vec{i_2}(t) \\
    \frac{d\vec{p_0}(t)}{dt} &= \vec{\lambda} \circ \vec{s}(t) - \omega_0 \vec{p_0}(t) \\
    \frac{d\vec{p_1}(t)}{dt} &= \omega_0 \vec{p_0}(t) - \omega_Y \vec{p_1}(t) \\
    \frac{d\vec{p_2}(t)}{dt} &= \omega_Y \vec{p_1}(t) - \omega_{X1} \vec{p_2}(t) \\
    \frac{d\vec{p_3}(t)}{dt} &= \pi \omega_{X1} \vec{p_2}(t) - \omega_{X2} \vec{p_3}(t)
\end{align}
where $\circ$ indicates element-wise multiplication.

@@

The mean prevalence on day $t$, used for the likelihood (see \cref{SEIR:eq:likelihood}), can be calculated as $\bar{p}_{at} = \frac{1}{2} \sum_{k=1}^l ( p_{kat} + p_{ka,t+1/2} )$.

@@

From the system of ODEs, the knowledge that all the compartments are growing exponentially at rate $\psi$, and the parameters $\vec{\pi}$ (where $\pi_a$ is the proportion of strata $i$ without immunity at time 0, hence, $\vec{r} = 1 - \vec{\pi}$) one can derive the following for the initial state of the compartments:
\begin{align}
    \vec{i_1} &= \vec{i_0} \left(1 + \frac{2\gamma}{\psi + 2\gamma} \right)^{-1} \\
    \vec{i_2} &= \vec{i_0} - \vec{i_1} \\
    \vec{e_2} &= \vec{i_1} \frac{\psi + 2\gamma}{2\sigma} \\
    \vec{e_1} &= \vec{e_2} \frac{\psi + 2\sigma}{2\sigma} \\
    \vec{s} &= \vec{\pi} - \vec{i_0} - \vec{e_1} - \vec{e_2} \\
    \vec{p_0} &= \vec{e_1} \frac{\psi + 2\sigma}{\psi + \omega_0} \\
    \vec{p_1} &= \vec{p_0} \frac{\omega_0}{\psi + \omega_Y} \\
    \vec{p_2} &= \vec{p_1} \frac{\omega_Y}{\psi + \omega_{X1}} \\
    \vec{p_3} &= \vec{p_2} \frac{\pi \omega_{X1}}{\psi + \omega_{X2}}.
\end{align}

@@

\ifSubfilesClassLoaded{
  \appendix
  \subfile{transmission-appendix}
  \listoftodos
}{}
